Simulating High-Dimensional Multivariate Data

### 2020-10-23 010-background.Rmd
decide on scope of RNA examples, revise introduction
revise introduction AGS

@@

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. For continuous marginals, the measure works well due to zero probability of ties. For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.

@@

>>>>

Formally, this is our target:

We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data. The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.

Ultra High Dimensional multivariate modeling and simulation desired properties, both basic properties (BP) and scaleability properties (SP). 
With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):

* BP1: Wide range of dependence, allowing both positive and negative dependence
* BP2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

Moreover, the simulation method must scale to high dimensions:

* SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

## Motivating example: RNA-seq data from breast cancer tumor samples

Simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018]. 
RNA-sequencing analysis has garnered much interest in the statistical literature (refs). 
Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). 
An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none'}
if ( !file.exists( 'data/brca99.rds' ) )  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-probs)*100% highest expressing genes for illustration
    myProb <- 0.99
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, 'data/brca99.rds')
}
```

```{r ch010-realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
brca <- readRDS(file = './data/brca99.rds')
set.seed(10192020)
exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
saveRDS(exampleGenes, file = './data/exampleGenes.rds')
kable(round(brca[1:4, exampleGenes]))
## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

Figure X.

```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
## Describe and plot real data
GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. For continuous marginals, the measure works well due to zero probability of ties.

@@

## Matching Pearson correlations

In the NORTA, it is well known that blah

## Discrete marginal considerations

<!-- 
For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.

@@

Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
Yet not all multivariate simulation approaches obtain these bounds [for example REF].
Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].

### 2020-10-23 index.Rmd
add some commented keywords
revise introduction AGS

--- 
title: Simulating High-Dimensional Multivariate Data using the bigsimr R Package
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
bibliography:
  - bigsimr.bib
  - packages.bib
site: bookdown::bookdown_site
journal: "Journal of Computational And Applied Mathematics (JCAM)"
#linenumbers: true
#numbersections: true
abstract: |
  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers often resort to simulation designs that posit independence. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This talk introduces the bigsimr R package that provides a general, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation matrices, and find close positive semi-definite matrices. Finally, we showcase the power of bigsimr by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
---

@@

```{r serveHelper, include=FALSE, eval=FALSE}
## rmote::start_rmote()
## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
## bookdown::preview_chapter("index.Rmd")
bookdown::serve_book()
```

@@

Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
Further efficient simulation empowers computational strategies (such as a parameteric bootstrap --- simulating from a hypothesized null model) to provide inference in analytically challenging settings.
Such Monte Carlo techniques become difficult for high-dimensional data with the current existing algorithms and tools.
This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study in this era of big data.

Along those lines, simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018]. 
RNA-sequencing analysis has garnered much interest in the statistical literature (refs). 
Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). 
An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.

## Ultra High Dimensional multivariate modeling and simulation desired properties

With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):

* P1: Wide range of dependence, allowing both positive and negative dependence
* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
* P3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

Moreover, the simulation method must scale to high dimensions:

* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
* S2: Procedure must scale to high dimensions while maintaining accuracy.

As others @MB13 have noted, however, simulating dependent data can be challenging. A central issue lies characterizing dependency between components in the high-dimensional random vector. The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g. non-normal, discrete, finite support, etc). Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics. For normal data, the Pearson product-moment correlation describes the dependency completely. As we will see, however, simulating arbitrary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**. On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application. Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. Yet not all multivariate simulation approaches obtain these bounds [for example REF]. Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].

To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula [refs]. The idea that many multivariate and marginally heterogeneous distribution can be constructed in such a manner is well known [refs]. This article's contribution lies in it's application to high-dimensional data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design. The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$. We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data. The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.

The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data. Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumptions --- emphasizing normal and non-normal scenarios. In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily available software. After evaluating *in silico*, we revisit our motivating
example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.

## Motivating example

- RNA-seq
- Provide a table of values
- check the talk history

```{r processBRCA, echo=FALSE, eval=TRUE, results='none'}
if ( !file.exists( 'data/brca99.rds' ) )  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-probs)*100% highest expressing genes for illustration
    myProb <- 0.99
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, 'data/brca99.rds')
}
```

```{r realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
brca <- readRDS(file = './data/brca99.rds')
set.seed(10192020)
exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
saveRDS(exampleGenes, file = './data/exampleGenes.rds')
kable(round(brca[1:4, exampleGenes]))
## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

Figure X.

```{r realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
## Describe and plot real data
GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - name: Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler
bibliography:
  - bigsimr.bib
  - packages.bib
site: bookdown::bookdown_site
#output: distill::distill_article
abstract: |
  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers often resort to simulation designs that posit independence. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This talk introduces the bigsimr R package that provides a general, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation matrices, and find close positive semi-definite matrices. Finally, we showcase the power of bigsimr by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
---

<!-- keywords simulation multivariate models high-dimensional data R packages -->

@@

```{r serveHelper, include=FALSE, eval=FALSE}
## rmote::start_rmote()
## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
## bookdown::render_book("index.Rmd", "bookdown::pdf_book", output_dir = "pdf")
## bookdown::preview_chapter("index.Rmd")
bookdown::serve_book()
```

@@

Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study in this era of big data.
As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
Further, efficient simulation empowers statistical computing strategies, such as the *parameteric bootstrap* [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
Such Monte Carlo techniques become difficult for high-dimensional data with the current existing algorithms and tools.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
For continous non--normal data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
Yet these approaches do not scale in a timely fashion to high-dimensional problems (see @Li2019gpu). 
For discrete data, multiple techniques have been proposed.
But some schemes have major flaws such as failing to obtain the full range of possible dependencies (for example, feasible for only positive correlations @Park1996). 
While more recent approaches [@MB13; @Xia17; @BF17] have largely rememdied this issue for low dimesional problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies characterizing dependency between components in the high-dimensional random vector.
The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g. non-normal, discrete, infinite support, etc). 
<!-- Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.  -->
For normal data, the Pearson product-moment correlation describes the dependency completely.
As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. 
On the other hand, an analyst may consider the use of non-parametric correlation measures to better characterize monotone dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
Throughout, we'll emphasize matching these nonparametric dependency measures as our aim lies in modeling non-normal data.

In the study, we present a scalable multivariate simulation algorithms and software. 
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software.
The algorithm design relies on useful properties of non-parameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.

The study proceeds with background and notation, including a description of our motivating example in RNA-sequencing (RNA-seq) breast cancer data.
Then we describe and justify our simulation methodology and related algorithms.
Next, a detailed illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package is provided.
Then we proceed with extensive Monte Carlo studies under various distributional assumptions.
After evaluating *in silico*, we revisit our high-dimensional motivating RNA-seq example and employ our methods to commonplace statistical computing tasks.
Finally, we'll make some concluding remarks regarding the method's utility/limitations and describe potential future directions.
